Transferability of clinical prediction models for early trauma care

### 2019-06-24 ManuscriptMarkdown.Rmd
Update ManuscriptMarkdown.Rmd

date: "June 18, 2019"

@@

output: word_document

@@

# Introduction
Trauma results in millions of deaths annually across the globe, and road traffic
injuries are among the top 10 leading causes of death worldwide [@RN1;
@RN21]. Each year 9% of the world’s deaths are because of trauma, and
predictions indicate that this number is likely to increase [@RN42].

In a typical, high resource setting, the initial management of patients is
performed on the scene of the trauma by Emergency Medical Services
(EMS). Patient data, vital signs and scene information is transferred to the
receiving hospital. This information is evaluated using a system to determine
the level of trauma and prepare adequate resources [@RN43], dictating if a full
or limited trauma team is activated [@RN44].

Systems to determine the level of trauma can be based on clinical prediction
models.  Models differ in quality and characteristics, but generally perform
well predicting survival [@RN45]. Many of these models are developed in a single
standardized setting such as a large metropolitan area or a major trauma centre
but are then implemented in different settings, such as non-trauma centres.

@@

# Materials and methods
## Study design
We conducted a registry-based cohort study. Sweden has a nationally encompassing
register for registering trauma called SweTrau. SweTrau data was used to create
clinical prediction models, which were transferred between different contexts to
study the effects on mistriage. The study and analysis plans were made
publically available before the research was undertaken [@RN59].

## Setting
In Sweden, trauma is the most common cause of death before 44 years of age. It
is estimated that up to 10% of all major trauma patients suffer some form of
disability. The societal effects of trauma are also substantial; with the loss
of work days comparable to those lost from cardiovascular and malignant diseases
combined [@RN23].

@@

Data was obtained from SweTrau between 2011 and 2016. Today 95,5% (52 of 55) of
Swedish hospitals record trauma cases in SweTrau. Currently the registry
includes 55 000 cases, and the board encourages its use in both academic
research and local quality improvement initiatives [@RN28].

The SweTrau inclusion criteria are: Traumatic event with subsequent activation
of hospital trauma protocol, admitted patients with NISS (New Injury Severity
Score) > 15 and patients transferred to the hospital within 7 days of traumatic
event with NISS > 15. SweTrau excludes patients if the only traumatic event is
chronic subdural hematoma or if hospital trauma protocol is activated without
traumatic event [@RN29].

## Participants 
The eligibility criteria were patients registered in SweTrau, age 15 or
above. This age was decided as the study aims to study adult trauma and not
paediatric trauma which differs in physiology, triage and initial care
[@RN46]. The Swedish guidelines for trauma activation define children as age <15
[@RN90]. Also, several guidelines and protocols listed in ATLS use age ≥15 as
cut-off [@RN43]. Patient age was obtained from the SweTrau register.

## Variables
### Model predictors
The clinical prediction models included the predictors systolic blood pressure
(SBP), respiratory rate (RR) and Glasgow coma scale (GCS) on arrival to
hospital. The rationale for including these three predictors is that they are
part of many established clinical prediction models for early trauma care, such
as the Revised Trauma Score [@RN35].

### Model outcome
The outcome used to develop the clinical prediction models was all cause
mortality within 30 days of the trauma.

@@

## Data sources and measurements
Model predictors, outcome, participant characteristics and study outcome as
outlined above were all obtained from SweTrau. The method of measurement for the
model predictors is not specified (for example, if SBP is measured using an
automated cuff or manually) in the registry entries. In Swedish emergency rooms,
patient parameters are usually obtained by a registered nurse or assistant nurse
and are assumed to be accurate.  Whether the patient is dead or alive 30 days
after trauma is manually entered into the registry locally, by each respective
hospital. Participant characteristics are obtained from the patient file before
being registered in SweTrau. NISS and ISS are calculated by hospital personal
based on the patient’s injuries.

## Bias
Data analysis was conducted according to a prearranged analysis plan. The
analysis plan and statistical analysis code was finalised using simulated
data. These efforts were taken to avoid confirmation bias. The analysis plan was
reviewed by an experienced statistician and programmer prior to implementation,
to ensure objectivity. Neither outcome nor variable were blinded when conducting
data analysis, which made a structured objective approach important.

@@

## Statistical methods
### Data sets
The complete SweTrau cohort was split into four sets of data. High and low
volume centres, metropolitan and non-metropolitan centres, multi and single
centre data and individual centres.

@@

### Development and validation sample
Each data set included at least two samples. The samples were then split into
two subsamples using a temporal split based on the date of trauma. The earlier
subsample was the development sample, and the later subsample the validation
sample. The development sample contained 70 events (events being patients who
died within 30 days of the trauma) and all non-events (non-events being patient
survival 30 days past the trauma) during the same time. The rationale for
including 70 events is that we needed at least 10 events per free parameter in
the logistic regression to obtain stable coefficient estimates [@RN30]. The
validation sample instead contained 100 events and at least 100 non-events,
which was suggested as the minimum number by Vergouwe in 2005 for external
validation samples [@RN37]. See figure 1 for example using the high- and low
volume centre data set, with high volume centres being those in the top quartile
of number of patients registered. The minimum sample size of development and
validation samples was 140 and 200 respectively. We performed analyses only on
data sets for which all samples included at least the minimum number of
patients.

@@

### Model development
In the model development step a clinical prediction model was developed in the
high-volume centre development sample. The model was developed using logistic
regression as implemented in the R function glm. The dependent variable was all
cause mortality within 30 days of trauma and independent variables SBP, RR, and
GCS modelled as previously described. To avoid overfitting the model a
bootstrap procedure was used to estimate a linear shrinkage factor that was then applied
to the model coefficients [@RN61]. The shrunk model was then used to estimate
the probability of all cause 30-day mortality in the development sample. A
gridsearch was performed across estimated probabilities in the development
sample to identify the cut-off that optimised overtriage keeping undertriage at
less than 5% [@RN39]. This cut-off was then used to classify patients as major
or minor trauma.

@@

### Missing data
We used multiple imputation using chained equations, as implemented in the R
package mice, to handle missing data [@RN41]. The number of imputations created
for each data set was equal to the highest percentage of missing data in that
data set. Quantitative variables were imputed using predictive mean matching and
qualitative variables were imputed using logistic regression. SBP and RR was
transformed as restricted cubic splines before imputation and imputed as just
another variable. All analyses outlined above were then conducted separately in
each imputed dataset. We present the combined results as the median point
estimate across imputations along with an empirical bootstrap of the 25th and
75th percentiles across imputations, i.e. the lower bound of the presented
interval is the lower bound of a 95% CI of the 25th percentile and the upper
bound is the upper bound of a 95% CI of the 75th percentile. This combined CI
was referred to as an Uncertainty Interval (UI) and was used to express the
added uncertainty associated with the imputation procedure to handle missing
data, as such it is more conservative than a standard 95% CI.

@@

**Table 1. Sample characteristics of multiple imputed data.** Data is presented as medians with interquartile range (IQR) or counts with % as applicable.
```{r, echo = FALSE, results = "asis"} 
## Table 1
options(knitr.kable.NA = '')
# kable(results$table.one.list$Characteristics_table_of_high.volume.vs.low.volume , format = "markdown")
# kable(results$table.one.list$Characteristics_table_of_metropolitan.vs.non.metropolitan , format = "markdown")
# kable(results$table.one.list$Characteristics_table_of_multi.centre.vs.single.centre , format = "markdown")

kable(table_one_high_low, format = "markdown")
kable(table_one_metro_non_metro, format = "markdown")
kable(table_one_multi_single, format = "markdown")
``` 
Abbreviations: GCS glasgow coma scale, SBP systolic blood pressure, RR respiratory rate, ISS injury severity score, NISS new injury severity score

@@

Table 2 shows model validation performance. The model with the lowest validation
mistriage rate was the `r colnames(lowest.val.mistriage)` model with a median
mistriage rate with 95% UI of `r lowest.val.mistriage`. Performance in terms of
undertriage and overtriage rate for the same model was `r val.undertriage`, and
`r val.overtriage` respectively. Worst validation performance i.e. highest
mistriage rate was found in the `r colnames(highest.val.mistriage)` sample, with
a model median mistriage rate of `r highest.val.mistriage`.

@@

Table 4 shows the transferred model performance when compared to the validation
performance in the sample reciving the transferred model, e.g. high volume
transfer performance minus low volume validation performance.

The highest mistriage rate when compared to the reciveing sample validation
model performance was found in the `r colnames(highest.comp.mistriage)` model,
with a median mistriage rate difference of `r highest.comp.mistriage`, meaning
this model performed worse when transferred to the other sample in the same data
set. In clinical terms, this model transfer means that among 100 trauma
patients, `r abs(worst.transfer$Transferred.mistriage.minus.buddy.local.mistriage.median) * 100` 
more patients would be wrongly classified as major or minor trauma.

@@

The most notable effects on model performance following model
transfer were seen after transferring the `r colnames(highest.comp.mistriage)`
model. The mistriage rate increased by `r substr(format(highest.comp.mistriage), 1, 5)` 
when using the `r colnames(highest.comp.mistriage)` model in the other
sample in the same data set. Mainly contributing to this increase in mistriage
was the marked increase in overtriage when performing this model transfer. For
this transfer the 95% UI ranged from `r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating 
a high level of uncertainty and that our findings are compatible with a `r UI.top.bot[1]` 
decrease to a `r UI.top.bot[2]` increase in mistriage. The negative 95% UI of `r UI.top.bot[1]` 
would signifiy an improvement in model performance. The higher
95% UI of `r UI.top.bot[2]` would mean a marked increase in model mistriage,
compared to the point estimate of `r substr(format(highest.comp.mistriage), 1, 5)`.

The transfer of clinical prediction models in trauma care has not previously
been extensively studied. In 2016, Gerdin et al. found model transfer to
adversely affect model performance in terms of calibration, but that this could
be improved by updating the model [@RN27]. This study did not examine model
performance in terms of calibration, neither was the effects of updating the
model explored.

External validation as the evaluation of a clinical prediction model in a
setting in which it was not originally developed has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when being externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62], showed models
significantly underestimating the probability of CAD. This study did not find
the same dramatic decrease in model performance after transfer. However, the
combined results of this study, the studies on model transfer and the studies on
external validation warrants further caution when transferring models between
different settings.

## Strengths and limitations
This study design realistically reflects potential transfers: e.g. clinical
prediction models for trauma are usually developed in large metropolitan centres
and then transferred as clinical recommendations to minor non-metropolitan
centres in which traumatic injuries are not as common.

Originally all single centres with a valid number of events were to constitute
one data set, but when performing the analysis only one centre provided a
sufficient number of events. Leading to the loss of the fourth data set:
Individual centres, and the comparison between these. Possibly, the results this
data set might have provided results that could reflect realistic situations as
major trauma centres regularly exchange data to improve patient care.

In Sweden major trauma is relatively uncommon, this might lead to
inconsistencies across centres when performing SweTrau registration. E.g. using
NISS instead of ISS, registering SBP by category and not by absolute value or
not following up on patient survival. This might lead to exaggerated performance
differences between the settings.

The models in this study used GCS, RR and SBP as predictors for 30-day
mortality. Potentially, the use of additional predictors such as age, mechanism
of injury, or newer predictors as shock index might increase model
performance. Furthermore, we used 30-day mortality as an outcome, recognizing
that late mortality, time in hospital, functional impairment, and morbidity are
also important outcome factors in the context of trauma care.

@@

## Clinical applications
The results of this study show that mistriage can potentially increase following
model transfer, in this case more prominently when transferring from a
metropolitan setting to a non-metropolitan setting. This supports the view that
model transfers should be done with caution, as performance might decline.

@@

Model transfer could potentially be studied using different methods. Quite
often, as mentioned, clinical models are transferred from a major metropolitan
centre to a smaller rural centre. Staff opinions considering how applicable this
new model is could be obtained using an interview or questionnaire approach
Clinical trials using different models could also be performed, however,
appropriate ethical considerations are paramount before performing this type of
study.

## Conclusion
Model transfer resulted in large and varying levels of uncertainty, and in all 
samples 95% UI included 0. Model transfer can therefore be unpredictable, but can 
potentially lead to increased mistriage, resulting in increased patient mortality 
and a misappropriation of resources, due to increased undertriage and overtriage, 
respectively. Transfer of the `r colnames(highest.comp.mistriage)` model lead to an 
increased mistriage rate of `r highest.comp.mistriage`. This observed effect was also
associated with considerable uncertainty, meaning mistriage might slightly
decrease, but would more likely increase, potentially substantially. Recognizing 
the limitations of this study, the authors believe further studies are warranted due
to the potential economic costs and patient consequences of poor model
transfers.

>>>>

date: "June 24, 2019"

@@

output:
  word_document:
    reference_docx: transferstyle.docx

@@

# Introduction
Trauma results in millions of deaths annually across the globe, and road traffic
injuries are among the top 10 leading causes of death [@RN1; @RN21]. 
Each year 9% of the world’s deaths are because of trauma, and
predictions indicate that this number is likely to increase [@RN42].

In a typical, high resource setting, the initial trauma management is
performed on the scene by Emergency Medical Services
(EMS). Patient data and vital signs are transferred to the receiving hospital. 
This information is then evaluated using a system to determine
the level of trauma, prepare adequate resources [@RN43] and dictate if a full
or limited trauma team is activated [@RN44].

Systems to determine the level of trauma can be based on clinical prediction
models. Models differ in quality and characteristics, but generally perform
well predicting survival [@RN45]. Many of these models are developed in a single
standardized setting such as a major trauma centre but are then implemented in 
different settings, such as non-trauma centres.

@@

# Materials and methods
## Study design
A registry-based cohort study was conducted. Sweden has a nationally encompassing
trauma register called SweTrau. SweTrau data was used to create
clinical prediction models, which were transferred between different contexts to
study the effects on mistriage. The study and analysis plans were made
publically available before the research was undertaken [@RN59].

## Setting
In Sweden, trauma is the most common cause of death before 44 years of age. Up 
to 10% of all major trauma patients suffer some form of disability. 
The societal effects of trauma are also substantial, with the loss
of work days comparable to those lost from cardiovascular and malignant diseases
combined [@RN23].

@@

Today 95,5% (52 of 55) of Swedish hospitals record trauma cases in SweTrau. 
Currently the registry includes 55 000 cases, and the board encourages its 
use in both academic research and local quality improvement initiatives [@RN28].

## Participants 
The eligibility criteria were patients registered in SweTrau between 2011 and 2016, 
age 15 or above. This age was decided as the study aims to study adult trauma and not
paediatric trauma which differs in physiology, triage and initial care
[@RN46]. Also, the Swedish guidelines for trauma activation define children as age <15
[@RN90]. Patient age was obtained from the SweTrau register.

## Variables
### Model predictors
The clinical prediction models used the predictors systolic blood pressure
(SBP), respiratory rate (RR) and Glasgow coma scale (GCS), all on arrival to
hospital. The rationale for these three predictors is that they are
part of many established clinical prediction models for early trauma care, such
as the Revised Trauma Score [@RN35].

### Model outcome
The outcome used to develop the clinical prediction models was all cause
mortality within 30 days of trauma.

@@

## Data sources and measurements
Model predictors, outcome, participant characteristics and study outcome as
outlined above were all obtained from SweTrau. The method of measurement for the
model predictors is not specified in the registry entries. In Swedish emergency rooms,
patient parameters are usually obtained by a registered nurse or assistant nurse
and are assumed to be accurate.  Whether the patient is dead or alive 30 days
after trauma is manually entered into the registry locally by each respective
hospital. NISS and ISS are calculated by hospital personal
based on the patient’s injuries.

## Bias
Data analysis was conducted according to a prearranged analysis plan. The
analysis plan and statistical analysis code was finalised using simulated
data. These efforts were taken to avoid confirmation bias. The analysis plan was
reviewed by an experienced statistician and programmer prior to implementation. 
Neither outcome nor variable were blinded when conductingdata analysis, which 
made a structured objective approach important.

@@

## Statistical methods
### Data sets
The complete SweTrau cohort was split into four sets of data, outlined below.

@@

### Development and validation sample
Each data set included at least two samples. The samples were then split into
two subsamples using a temporal split based on the date of trauma. The earlier
subsample was the development sample, and the later subsample the validation
sample. The development sample contained 70 events (events being patients who
died within 30 days of the trauma) and all non-events (non-events being patient
survival 30 days past the trauma) during the same time. The rationale for
including 70 events is that at least 10 events per free parameter were needed in
the logistic regression to obtain stable coefficient estimates [@RN30]. The
validation sample instead contained 100 events and at least 100 non-events,
which was suggested as the minimum number by Vergouwe in 2005 for external
validation samples [@RN37]. See figure 1 for example using the high- and low
volume centre data set. The minimum sample size of development and
validation samples was 140 and 200 respectively. We performed analyses only on
data sets for which all samples included at least the minimum number of
patients.

@@

### Model development
In the model development step a clinical prediction model was developed in the
high-volume centre development sample. The model was developed using logistic
regression. The dependent variable was all cause mortality within 30 days of 
trauma and independent variables SBP, RR, andGCS modelled as previously described. 
To avoid overfitting the model a bootstrap procedure was used to estimate a linear 
shrinkage factor that was then appliedto the model coefficients [@RN61]. The shrunk 
model was then used to estimate the probability of all cause 30-day mortality in the 
development sample. A gridsearch was performed across estimated probabilities in 
the development sample to identify the cut-off that optimised overtriage keeping 
undertriage at less than 5% [@RN39]. This cut-off was then used to classify patients 
as major or minor trauma.

@@

### Missing data
We used multiple imputation using chained equations (mice), to handle missing data [@RN41]. 
The number of imputations created
for each data set was equal to the highest percentage of missing data in that
data set. Quantitative variables were imputed using predictive mean matching and
qualitative variables were imputed using logistic regression. SBP and RR was
transformed as restricted cubic splines before imputation. All analyses outlined 
above were then conducted separately in
each imputed dataset. We present the combined results as the median point
estimate across imputations along with an empirical bootstrap of the 25th and
75th percentiles across imputations, i.e. the lower bound of the presented
interval is the lower bound of a 95% CI of the 25th percentile and the upper
bound is the upper bound of a 95% CI of the 75th percentile. This combined CI
was referred to as an Uncertainty Interval (UI) and was used to express the
added uncertainty associated with the imputation procedure to handle missing
data, as such it is more conservative than a standard 95% CI.

@@

**Table 1. Sample characteristics of multiple imputed data.** Data is presented as medians with interquartile range (IQR) or counts with % as applicable.
```{r, echo = FALSE, results = "asis"} 
## Table 1
options(knitr.kable.NA = '')
# kable(results$table.one.list$Characteristics_table_of_high.volume.vs.low.volume , format = "markdown")
# kable(results$table.one.list$Characteristics_table_of_metropolitan.vs.non.metropolitan , format = "markdown")
# kable(results$table.one.list$Characteristics_table_of_multi.centre.vs.single.centre , format = "markdown")
kable(table_one_high_low, format = "markdown")
kable(table_one_metro_non_metro, format = "markdown")
kable(table_one_multi_single, format = "markdown")
``` 
Abbreviations: GCS glasgow coma scale, SBP systolic blood pressure, RR respiratory rate, ISS injury severity score, NISS new injury severity score

@@

Table 2 shows model validation performance. The model with the lowest validation
mistriage rate was the `r colnames(lowest.val.mistriage)` model with a median
mistriage rate of `r lowest.val.mistriage`. Performance in terms of
undertriage and overtriage rate for the same model was `r val.undertriage`, and
`r val.overtriage` respectively. Worst validation performance i.e. highest
mistriage rate was found in the `r colnames(highest.val.mistriage)` sample, with
a model median mistriage rate of `r highest.val.mistriage`.

@@

Table 4 shows the transferred model performance when compared to the validation
performance in the sample reciving the transferred model.
The highest mistriage rate when compared to the reciveing sample validation
model performance was found in the `r colnames(highest.comp.mistriage)` model,
with a median mistriage rate difference of `r highest.comp.mistriage`, meaning
this model performed worse when transferred to the other sample in the same data
set. In clinical terms, this model transfer means that among 100 trauma
patients, `r abs(worst.transfer$Transferred.mistriage.minus.buddy.local.mistriage.median) * 100` 
more patients would be wrongly classified as major or minor trauma.

@@

The most notable effects on model performance following model
transfer were seen after transferring the `r colnames(highest.comp.mistriage)`
model, with a mistriage rate increase by `r substr(format(highest.comp.mistriage), 1, 5)`. 
Mainly contributing to this was a marked increase in overtriage. 
For this transfer the 95% UI ranged from `r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating 
a high level of uncertainty and that our findings are compatible with a `r UI.top.bot[1]` 
decrease to a `r UI.top.bot[2]` increase in mistriage. The negative 95% UI of `r UI.top.bot[1]` 
would signifiy an improvement in model performance. The higher
95% UI of `r UI.top.bot[2]` would mean a marked increase in model mistriage,
compared to the point estimate of `r substr(format(highest.comp.mistriage), 1, 5)`.

The transfer of clinical prediction models in trauma care has not previously
been studied extensively . In 2016, Gerdin et al. found model transfer to
adversely affect model performance in terms of calibration, but that this could
be improved by updating the model [@RN27]. This study did not examine model
performance in terms of calibration, neither were the effects of updating the
model explored.

External validation as the evaluation of a clinical prediction model in a
setting in which it was not originally developed has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when being externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62], showed models
significantly underestimating the probability of CAD. The
combined results of this study, the studies on model transfer and the studies on
external validation suggests caution when transferring models.

## Strengths and limitations
This study design realistically reflects potential transfers: clinical
prediction models for trauma are usually developed in large metropolitan centres
and then transferred as clinical recommendations to minor non-metropolitan
centres.

Originally all single centres with a valid number of events were to constitute
a data set, but only one centre provided a sufficient number of events. Leading to 
the loss of the data set: Individual centres. This data set might have provided
results that could reflect realistic situations as major trauma centres regularly 
exchange data.

This study used GCS, RR and SBP as predictors for 30-day
mortality. Additional predictors such as age, mechanism
of injury, or newer predictors as shock index might increase model
performance. Furthermore, we used 30-day mortality as outcome, recognizing
that late mortality, time in hospital, functional impairment, and morbidity are
also important outcome factors in the context of trauma care.

@@

## Clinical applications
This study shows that mistriage can potentially increase following
model transfer, and more prominently in certain cases. This supports 
the view that model transfers should be done with caution, as 
performance might decline.

@@

Model transfer could potentially be studied using different methods. Staff opinions 
considering how applicable this new model is could be obtained using a questionnaire approach. 
Clinical trials using different models could also be performed, however, appropriate ethical 
considerations are paramount before performing this type of study.

## Conclusion
Model transfer resulted in large and varying levels of uncertainty. Model transfer 
can therefore be unpredictable, and potentially lead to increased mistriage, resulting 
in increased patient mortality and a misappropriation of resources, due to increased 
undertriage and overtriage, respectively. Transfer of the `r colnames(highest.comp.mistriage)` 
model lead to an increased mistriage rate of `r highest.comp.mistriage`. This 
observed effect was also associated with considerable uncertainty, meaning mistriage might slightly
decrease, but would more likely increase, potentially substantially. Recognizing 
the limitations of this study, the authors believe further studies are warranted due
to the potential economic costs and patient consequences of poor model
transfers.
